Algebraic Groups\vspace{-1ex}

### 2020-01-29 Algebraic Geometry/Algebraic-Geometry.tex
update

\subsection{Examples}
Consider $f\in k[x,y]$. THen $\Spec k[x,y]/(f)$ is integral.
Consider $n$ copies of $\A1$ glued along $\A1\setminus 0$. This is also integral.

@@

>>>>

\subsection{Examples}
Consider $f\in k[x,y]$. Then $\Spec k[x,y]/(f)$ is integral.
Consider $n$ copies of $\A1$ glued along $\A1\setminus 0$. This is also integral.

@@

\section{January 29th, 2020}
\subsection{Example}
Consdier the tangent space (the $\Hom$ from $\Spec k[\varepsilon]/(\varepsilon^2)$) to the space $\Spec k[x,y]/(xy)$ (the axes in $\bbR^2$).

Looking at the algebra maps, you can map $x$ and $y$ freely to multiples of $\varepsilon$, giving us a two dimensional thing! Otherwise you can compute you only have one 
degree of freedom. It is (locally) Noetherian, reducible, integral, and other stuff.

You can do this with higher-order vectors as well! Think about what it means to map $\Spec k[\varepsilon]/(\varepsilon^n)$ into a scheme.

Think more about the countable number of lines glued together in a criss-cross. This is a scheme! It is locally Noetherian but not Noetherian. Also think about the line with infinitely many origins.

\subsection{AG is relative}
\begin{defn}
	A \textbf{morphism $f:X\to Y$ of schemes} is 
	\begin{itemize}
		\item \textbf{quasicompact} if there is an open cover $U_i\subseteq Y$ with $U_i$ and $f^{-1}(U_i)$ quasicompact.
		\item \textbf{locally of finite type} if there exists an open cover $\Spec A_i=U_i\subseteq Y$ such that each $f^{-1}(U_i)$ has an open cover $\Spec B_{ij}$ such that each map $A_i\to B_{ij}$ 
		makes $B_{ij}$ into a finitely generated $A_i$ algebra.
		\item \textbf{of finite type} if $f$ is locally of finite type and quasicompact.
		\item \textbf{affine} if there exists an open covering $\Spec A\subseteq Y$ such that $f^{-1}(U_i)$ is affine.
		\item \textbf{finite} if it is affine and there is an open cover $\Spec A_i\subseteq Y$ such that $f^{-1}(U_i)=\Spec B_i\to \Spec A_i$ is (module-)finite.
		\item \textbf{quasifinite} if the map of topological spaces has finite fibers.
		\item \textbf{locally quasifinite} if there exists a cover $U_i\subseteq Y$ and $V_{ij}\subseteq X$ such that $V_{ij}\to U_i$ is quasifinite.
	\end{itemize}
\end{defn}

This is a theorem with a thousand different formulations so you may not find this one online:
\begin{thm}[Zariski's Main Theorem]
	Finite morphisms are precisely the proper and quasifinite ones.
\end{thm}

\subsection{An observation}
Let $X$ a variety. Then how do we understand this in the language of schemes? It is an \textit{integral} $k$-scheme such that the map 
$X\to \Spec k$ is
\begin{itemize}
	\item of finite type,
	\item and separated (will see this later).
\end{itemize}
So finite type is aprt of ``relative variety.'' We can imagine a family of varieties given by the zero sets 
of $y^^2=x^3+tx+4$ and consider this as a closed subvariety in $\A 2$ over $\A1$. Then we could specialize at any prime and recover a variety in $\A2_\bbZ$.
Furthermore varying $4$ can get us more.

\subsection{Another observation}
If $f:X\to Y$ is finite, then $f$ is quasifinite. To see this, let $y\in \Spec A\subset Y$. Then since $f$ is finite,
$f^{-1}(\Spec A)=\Spec B$ and $A\to B$ is a finite ring map (some ordering needs to be fixed there on choosing $A$ and $B$).
Thus without loss of generality, we can assume $X=\Spec A$ and $Y=\Spec B$. So then $y\in Y$ if and only if $p\lhd A$ is prime.

### 2020-01-29 Algebraic Groups/Algebraic-Groups.tex
update

>>>>

\section{January 29th, 2020}
As a reminder: we are not having classes February 10-14. The following Monday is president's day. So keep that in mind. :)

\subsection{Lie algebras for \texorpdfstring{$G$}{G}}
\begin{defn}
	If $G$ is an algebraic group scheme over $k$, then $\Lie G$ is the space of left-invariant derivations $D:k[G]\to k[G]$.
\end{defn}
What is a left-invariant derivation? Let $A=k[G]=k^G$. The Yoneda lemma allows us to identify this with $\Hom(G,\A1)$.
But given $f:G\to k$, we get an action 
\[g\cdot f(-)=(T_g f)(-)=f(g-)\]
Note that this actually defines a \textit{right action}, but the left and right invariant derivations are canonically isomorphic and this 
will help us with the Lie algebra case.

Now let $x\in G$ and by abuse of notation let $x=\operatorname{ev}_x:A\to k$ be the evaluation map. This gives us a diagram:
\begin{figure}[h]
	\begin{tikzcd}
		A\ar[r,"\Delta"]\ar[d,"T_g"] & A\otimes A\ar[r,"(g;\operatorname{id})"] & A\ar[d,"x"]\\
		A\ar[rr,"x"] & & k
	\end{tikzcd}
\end{figure}

Let $T:A\to A$ be a $k$ map. THen $T$ is left invariant if $T_g\circ T=T\circ T_g$ for all $G$. 
I wrote some notes in my digital pad. We did some computation. The result is that it is equivalent to having $\Delta\circ T=(\id\otimes T\circ \Delta)$.

Thus we can restate the definition of $\Lie G$ as 
\[\Lie G=\{D\in\Der_k(A,A)|\Delta\circ D=(\id\otimes D)\circ \Delta\}.\]
\begin{lem}
	$\Lie G$ has the structure of a Lie algebra with bracket 
	\[[D_1,D_2]=D_1\circ D_2-D_2\circ D_1\]
\end{lem}
To prove that this we need to show the three things for the Lie bracket hold! The $k$ bilinearity and $[D,D]=0$ (careful! We want to consider characteristic 2!)
are easy to show but the Jacobi identity requires some work.

Notice that it is equivalent to saying that the operator $\ad D$ is a derivation:
\[\ad D([D_1,D_2])=[\ad D(D_1),D_2]+[D_1,\ad D(D_2)]\]
and one can check this. It has nothing to do with left invariance!

\subsection{Another way to think of this algebra}
\begin{lem}
	There are natural bijections between 
	\begin{enumerate}
		\item $\Lie G$ (the left invariant derivations on $k[G]$)
		\item $\Der_\varepsilon(k[G],k)$
		\item $\ker (G(k[\tau]/(\tau^2))\to G(k))$ where the map is induced from the projection onto the constants in $k[\tau]/(\tau^2)$.
	\end{enumerate}
\end{lem}
\begin{prf}
	We can use that $G(k[\tau]/(\tau^2))=\Hom(A,k\oplus k\tau$ and then study the kernel.
\end{prf}
